A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950, including the Stensele station and the inland

### 2022-02-04 paper3.Rmd
more on B-P and polish

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, global.par=TRUE)
library(knitr)
library(eha)
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

@@

@@

Temperature data are collected from two weather stations, *Umeå* and *Bjuröklubb*
(used with population data from Skellefteå coastal area). Both stations deliver
daily temperature data covering our time period, usually three measures per day, morning,
noon, and evening. In Table \@ref(tab:typweek), the Umeå data from the week 1--7 January,
1923 is shown.

@@

```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 3}
oldpar <- par(mfrow = c(1, 2))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
par(oldpar)
```

@@

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[neo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanals, results = 'asis'}
##library(survival)
summer <- neo[neo$season %in% c("summer", "fall"), ]
summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = summer)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

```{r getpneo}
pneo <- age.window(infdat, c(28 / 365, 1))
gpneo <- pneo
gpneo$enter <- g(gpneo$enter * 365)
gpneo$exit <- g(gpneo$exit * 365)
##
gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
                    lowTemp + highTemp + season + aver
                + emeantemp + subreg, 
                data = gpneo, cuts = g(c(28, 365)))
```

@@

```{r nap, fig.cap = "Nelson-Aalen estimator and exponential cumulative hazards of postneonatal survival on the g scale.",fig.height=3.2}
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = gpneo)
plot(fit, xlam = "g(t)")
fit.w <- tpchreg(oe(event, exposure) ~ 1, data = gtpneo, time = age)
zz <- hazards(fit.w)
lines(zz$x, zz$y, col = "red", lty = 2)
legend("topleft", legend = c("Nelson-Aalen", "Exponential"), lty = 1:2, col = c("blue", "red"))
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + socStatus  + 
                        illeg + parity + subreg, 
                    data = winter, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + socStatus  + 
                        illeg + parity + subreg, 
                    data = summer, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA)
knitr::opts_knit$set(global.par = TRUE)
##
library(knitr)
library(eha)
##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
g <- function(x) (log(1 + x))^3 # x in days.
dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
ginv <- function(y) exp(y^(1/3)) - 1
```

```{r setpar}
par(las = 1, cex.axis = 0.8, cex.lab = 0.8, cex = 0.8)
```

@@

Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
marked. The map is taken from the paper by @junkka.

```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
include_graphics("myfigs/skum.png")
```

@@

Temperature data are collected from three weather stations, *Umeå*, *Bjuröklubb*
(used with population data from Skellefteå coastal area), and *Stensele* (Inland). All stations deliver
daily temperature data covering our time period, usually three measures per day, morning,
noon, and evening. In Table \@ref(tab:typweek), the Umeå data from the week 1--7 January,
1923 is shown.

@@

```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3}
oldpar <- par(mfrow = c(1, 3))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
umeanavg <- umetemp$emeantemp[1:52]
##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
plot(1:52, umaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
m12 <- c(-30, -15, 0, 14.5, 25)
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, umeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
bmeanavg <- bjurtemp$emeantemp[1:52]
plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
## Inland:
bmaxavg <- stentemp$emaxtemp[1:52]
bminavg <- stentemp$emintemp[1:52]
bmeanavg <- stentemp$emeantemp[1:52]
plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Inland", cex = 0.7)
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1, at = m12)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = m12, lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
text(8, 11, "max", col = "red", cex = 0.7)
text(20, -9, "min", col = "blue", cex = 0.7)
box()
##
par(oldpar)
```

@@

```{r neoanalw, results = 'asis'}
##library(survival)
winter <- neo[neo$season %in% c("winter", "spring"), ]
winter$season <- factor(winter$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = winter)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = 3)
```

@@

```{r neoanals, results = 'asis'}
##library(survival)
summer <- neo[neo$season %in% c("summer", "fall"), ]
summer$season <- factor(summer$season)
##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
##neo$enter <- g(neo$enter * 365)
fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
                     sex + season + subreg + illeg + parity + socBranch,
                 data = summer)
dr <- drop1(fit.cr, test = "Chisq")
ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = 2)
```

@@

```{r getpneo}
pneo <- age.window(infdat, c(28 / 365, 1))
gpneo <- pneo
gpneo$enter <- g(gpneo$enter * 365)
gpneo$exit <- g(gpneo$exit * 365)
##
oj <- system.time(gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + illeg + parity + 
                    lowTemp + highTemp + season + aver
                + emeantemp + subreg, 
                data = gpneo, cuts = g(c(28, 365))))
```

@@

```{r nap, fig.cap = "Nelson-Aalen estimator and exponential cumulative hazards of postneonatal survival on the g scale.",fig.height=3.2}
par(cex = 0.8)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = gpneo)
xx <- hazards(fit)
plot(xx[[1]][, 1], cumsum(xx[[1]][, 2]), type = "s", xlab = "Age (days)",
     col = "blue", ylab = "Cumulative hazards", axes = FALSE)
axis(1, at = g(c(28, 183, 365)), labels = c(28, 183, 365))
axis(2)
box()
fit.w <- tpchreg(oe(event, exposure) ~ 1, data = gtpneo, time = age)
zz <- hazards(fit.w)
lines(zz$x, zz$y, col = "red", lty = 2)
legend("topleft", legend = c("Nelson-Aalen", "Exponential"), lty = 1:2, col = c("blue", "red"))
```

The survival data set we work with here contains 102 thousand infants and 4.9 million records,
an average of 48 records per infant, one record per calendar week in life. In total there are 
`r sum(gpneo$event)` deaths, and a typical Cox regression execution time is slightly
below one minute. A "naive" application of an exponential regression model to the same data
is typically using around six minutes.

However, with the exponential model (or generally, the *piecewise constant hazard* (pch)) in mind, data can be heavily reduced by *tabulation*, and the
statistical *sufficiency principle* will guarantee that we get identical results.
The tabulation itself takes about 20 seconds, and the analysis after that about 15 seconds, a
total of 35 seconds. But the tabulation usually only needs to be done once. More about this later.

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = winter, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = 4)
```

@@

fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                        illeg + parity + subreg, 
                    data = summer, time = age)
dr <- drop1(fit.tpch, test = "Chisq")
ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = 4)
```

@@

```{r bpprim, fig.cap = "Postneonatal survival is Weibull vs. Nelson-Aalen.", echo = TRUE}
fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y, col = "red", lty = 2)
```

Not very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
*exogeneous infant mortality* is Weibull distributed. Look here (Figure \@ref(fig:ltruw)). A perfect fit, 
just like the one under the $g$- transform. This is the way to go, *more will come tomorrow*!

```{r ltruw, fig.cap = "Postneonatal survival is truncated Weibull vs. Nelson-Aalen.", echo = TRUE}
fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
plot(fit, fn = "cum", col = "blue")
xx <- hazards(fit.w)
lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
```
